---
title: "Crash Score"
author: "Minh Ho"
date: "8/8/2021"
output: html_document
---
```{r}
df <- read.csv("/Users/apple/R studying/Crash Score project/2019-06-13-exam-pa-data-file.csv")
# Load library
library(ggplot2)
library(dplyr)
library(plyr)
library(caret)
```

```{r}
summary(df)
```
```{r}
summary(as.factor(df$Time_of_Day))
df1 <- df
df1$Time_of_day <- as.factor(df1$Time_of_Day)
ggplot(data = df1, aes(x = Time_of_day)) + geom_bar()
```

Check to see if any missing value

```{r}
for (i in names(df)) {
  print(i)
  print(sum(is.na(df[, i])))
}

#Other way to do that
apply(is.na(df), 2, sum)
```
We can see that there is no missing value for any variables. That would be great. I would then draw some charts to see dig deeper into the variables

```{r}

factor_list <- c()
others <- c()
for (i in names(df)) {
  if (class(df[ , i]) == "factor") {
    factor_list <- append(factor_list, i)
  }
  else {
    others <- append(others, i)
  }
}

for (i in factor_list) {
  p <- ggplot(data = df, aes(x = df[, i])) + 
    geom_bar() + 
    xlab(i) + 
    theme(axis.text.x = element_text(angle = 12),
          panel.background = element_blank(),
          panel.grid = element_line(colour = "grey"))
  print(p)
}

for (i in others) {
  p <- ggplot(data = df1, aes(x = df1[, i])) + geom_histogram() + xlab(i)
  print(p)
}
```
the most frequent road feature is NONE
the most frequent Road Character is Straight level, next is Straight Grade. Others characters are very rare.
Rd_Class is half high way, half other.
Rd_Configuration is mostly two way no median.
Rd_surface are mostly smooth asphalt 
Rd_conditions are mostly dry, sometimes is wet and very rare it is ice-snow-slush and other
Light is mostly daylight. sometimes is dark-lit, very rare it is dark-not-lit, dawn, dusk and other.
Most of the time it is clear, sometimes cloudy, rain, snow and other is very rare.
Traffic control most of the time it is Non, signal is also frequent, sometimes it is stop-sign, yield and other is very rare.
Most of the time the accident happens in non-work-area.
Crash_Score is skew left.
Year seems to be increasing every year, except 2019, seems to be the time of record is at that middle of 2019.

#RELATIONSHIP BETWEEN TARGET AND PREDICTORS

I believe that the year, month should be factor. Therefore I convert them into factor before doing this analysis.

```{r}
df1 <-df
df1$year <-as.factor(df1$year)
df1$Month <- as.factor(df1$Month)
factor_list <- c()
others <- c()
for (i in names(df1)) {
  if (class(df[ , i]) == "factor") {
    factor_list <- append(factor_list, i)
  }
  else {
    others <- append(others, i)
  }
}

for (i in factor_list) {
  p <- ggplot(data = df1, aes(x = df1[, i], y = Crash_Score)) + 
    geom_boxplot() + 
    xlab(i) + 
    ylim(0, 15) +
    theme(axis.text.x = element_text(angle = 12)
          )
  print(p)
}

for (i in others) {
  p <- ggplot(data = df1, aes(x = df[, i], y = Crash_Score)) + geom_point() + xlab(i)
  print(p)
}
```
Actually I think there is a lot of outliers in the boxplot Crash_Score and other features. I also see that the Time_of_day should be factor as well. I modify a bit on my code. First I would like to make Crash_Score related feature being normal first by log transforming it

```{r}
ggplot(data = df, aes(x = log(Crash_Score))) + geom_histogram()
```
it looks so much normal than before. Then I copy the code before, making time of day being factor and then see the result.

```{r}
df2 <- df
df2$year <-as.factor(df2$year)
df2$Month <- as.factor(df2$Month)
df2$Time_of_Day <- as.factor(df2$Time_of_Day)
df2$ln_Crash_Score <- log(df2$Crash_Score)
df2$Crash_Score <- NULL

factor_list <- c()
others <- c()
for (i in names(df2)) {
  if (class(df2[ , i]) == "factor") {
    factor_list <- append(factor_list, i)
  }
  else {
    others <- append(others, i)
  }
}

for (i in factor_list) {
  p <- ggplot(data = df2, aes(x = df2[, i], y = ln_Crash_Score)) + 
    geom_boxplot() + 
    xlab(i) + 
    ylim(0, 4) +
    theme(axis.text.x = element_text(angle = 12)
          )
  print(p)
}


```
We can see that when we log transformed the Crash Score value, it is better to see the boxplots. 
For year, month I don't see any differences in the predictor value and the ln_Crash_Score.
For Rd_feature, Intersection seems to make the Rd_feature being a bit higher in the ln_Crash_score
For Rd_Character, Other has lower ln_Crash_Score. Straight seems to have higher ln_Crash_score than Curve.
HWY have higher Crash_Score than Other. 

#REDUCE THE NUMBER OF LEVELS 
I have to look at each factor variable and see if I can group the factor levels together if the outcome in the target is the same

```{r}
for (i in names(df)) {
  table <- df%>%group_by_(i) %>% summarise(mean = mean(Crash_Score),
                                           median = median(Crash_Score),
                                           n = n())
  print(table)
}

```

Comments:
Years, Months seem not related to the Crash_Score we want to explore.
Rd_Feature make some difference between levels. I decide to keep it untouched
Rd_Class has significant difference enough 
Rd_Configuration has the Unknown level very small amount of observation, and its mean and median is close to One-Way, so I think I would merge the two levels into One-Way
Rd_Surface doesn't have any mergeable levels as each level differs.
Rd_Condition, Light and Weather will not be touched as asked.
Traffic_Control has a clear different between levels: None < Yield < Other < Stop-sign < Signal. I would not change any level here as it is very clear.
Work_Area is clear enough not to touch.

```{r}
# Merge level One-Way and Unknown to 1 level

var <- "Rd_Configuration"
df[,var] <- as.factor(df[,var])
var.levels <- levels(df[,var])
df[,var] <- mapvalues(df[,var],var.levels,c("ONE-WAY", "TWO-WAY-NO-MEDIAN", "TWO-WAY-PROTECTED-MEDIAN", "TWO-WAY-UNPROTECTED-MEDIAN", "ONE-WAY"))

# We can change the "Rd_Configuration" to var to get the same result
table <- as.data.frame(table(df[,"Rd_Configuration"]))
  max <- which.max(table[,2])
  level.name <- as.character(table[max,1])
  df[,"Rd_Configuration"] <- relevel(df[,"Rd_Configuration"], ref = level.name)
table(df[,"Rd_Configuration"])
```
# Running Principal Component Analysis

```{r}
dfPCA <- df[ , -which(names(df) == "Crash_Score")]
mainEffects <- dummyVars(~., data = dfPCA)
dfPCA_bin <- predict(mainEffects, newdata = dfPCA)
PCA <- prcomp(dfPCA_bin, center = TRUE, scale = TRUE)
summary(PCA)
```

We have 50 PC in total but just first 25 PCs has explained more than 80% of the data. If I need to just take a few PCs to get the reduction of dimension I would do this way. However, this method make the model harder to explain, that's the drawback.

If I have to reduce the number of dimension using PC, I would seriously consider this option. However my main goal here is to predict the Crash_Score, so I would consider using PCA to generate a new feature from previous variables.

Now I will use the advice from the Solution (only use 3 variables and then extract the Principal Component(s)) (only Rd_conditions, Light and Weather)
```{r}
dfPCA1 <- df[ , c("Rd_Conditions", "Light", "Weather")]
# use dummyVars function
mainEffects <- dummyVars(~., data = dfPCA1)
dfPCA1_bin <- predict(mainEffects, newdata = dfPCA1)
PCA1 <- prcomp(dfPCA1_bin, center = TRUE, scale = TRUE)
summary(PCA1)
```

```{r}
PCA1$rotation
```

We can see that the first PC the strongest weight will be:
Rd_conditions.DRY 


